Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2016-01-21 ijcar2016/body.tex
start writing section about Nunchaku

%%% BEGIN Times font
\usepackage{mathptmx}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
%%% END Times font

@@

\newcommand\closurefmla{\psi \models_T \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}}

@@

  * Nunchaku
    * intended as a higher-order model finder with a rich input language,
      including (co)datatypes, (co)recursive functions, and
      (co)inductive predicates
  * first version, 0.1, came out in January 2016, without support for
    higher-order constructs yet
  * standalone tool; have a frontend for Isabelle and are envisioning
    frontends for other proof assistants, notably Coq and the
    TLA\textsuperscript{+} proof system
  * displace Nitpick, based on Kodkod
    * CVC4 has built-in (co)datatypes and efficient arithmetic
    * also, Nitpick is intimately tied to Isabelle

  * advantage over Nitpick:
    * generates a single problem, not parameterized by cardinalities
      * pushes all the work to the SMT solver
        * we are just starting with that -- and see lots of potential for
          optimizations and tuning

  * flavor of input language
    * case expressions

  * architecture: pipeline of transformations, each of which consists of a
    pair of functions $(translate, reconstruct)$:
    * translates transforms a problem over a signature $\Sigma$ in a logic $\mathcal{L}$
      to a problem over a signature $\Sigma'$ in a logic $\mathcal{L}'$
    * reconstruct transforms a model back from $\Sigma'$ to $\Sigma$
    * similar to institutions \cite{goguen-meseguer}

  * table with the current pipeline

  * rec transformation is just one operation in pipeline
    * reconstruct function: transforms a total model of the encoded problem
      into a partial model of the original model, trimming out junk and adding
      "..." to represent missing entries

  * higher-order support will be done as a natural extension of the encoding
    presented in this paper
    * explain a bit
    * two o

>>>>

%%% BEGIN Times font
\usepackage{mathptmx}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}

%%% END Times font

@@

\newcommand\closurefmla{
  \ensuremath{\psi \models_T \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}} }

@@

Nunchaku is a new model finder, intended
to be used for finding counter-examples in several proof assistants. The
first version, 0.1, was released in January 2016 with support for (co)datatypes,
(co)recursive functions and (co)inductive predicates.
Support for higher-order
functions is planned in the next release.
We intend for Nunchaku to be supported at least in Isabelle (where basic
support for it already exists), Coq, and the TLA\textsuperscript{+} proof
system.

Nunchaku is the spiritual successor to Nitpick~\cite{blanchette-nipkow-2010},
but is developed as a standalone OCaml program (with its own input language) so
as to be usable independently from Isabelle.
Whereas Nitpick relies on propositional checking with Kodkod,
generating a succession of problems where cardinalities of finite types
grow at each step, Nunchaku
translates its input to one first-order logic formula that targets
the finite-model finding fragment of CVC4, including the (co)datatypes support.
Using CVC4 will also allow Nunchaku to provide efficient arithmetic and makes
equality reasoning straightforward.

The following problem is expressed in Nunchaku's input syntax; it starts with
a datatype declaration, then two mutually recursive coinductive predicates
are defined, and finally a goal (``find a non-zero natural number $m$ that
is even'').
\begin{quote}
\begin{verbatim}
data nat := zero | Suc nat.

copred even : nat -> prop :=
  even zero;
  forall (n : nat). odd n => even (Suc n)
and odd : nat -> prop :=
  forall (n : nat). even n => odd (Suc n).

goal exists m. (even m && ~ (m = zero)).
\end{verbatim}
\end{quote}

Nunchaku parses and types the input problem before applying a sequence
of translations, each reducing the distance to the target fragment. Here,
for instance, the coinductive predicates
\verb!even! and \verb!odd! are \emph{polarized} (specialized into
\verb!even+!, used only in positive positions, and \verb!even-! used in negative
positions),
$m$ is Skolemized into a fresh constants,
then then translated into admissible functions
before another pass uses the encoding from Section~\ref{sec:encoding}.
If a model is found, it is translated back to the input language.
Conceptually, the sequence of transformation is a \emph{two ways pipeline}
built by composing pairs \textsf{(encode,decode)}.
For each such pair, \textsf{encode}
transforms a problem over a signature $\Sigma$ in a logic
$\mathcal{L}$ to a problem over a signature $\Sigma'$ in a slightly different
logic $\mathcal{L}'$, and \textsf{decode} translates a model of \mathcal{L'}
over $\Sigma'$ into a model of $\mathcal{L}$ over $\Sigma$. The table
from Figure~\ref{fig:pipeline} presents the current pipeline implemented in Nunchaku.

\begin{figure}[tb]
%{{{
\begin{tabular}{|l|p{0.6\textwidth}|}
  \hline
  component & description \\
  \hline
  \hline
  Type inference
    & Infer types and checks definitions \\
  \hline
  Type skolemization
    & Replaces $\exists \alpha. p[\alpha]$
      with $p[\tau]$ where $\tau$ is a fresh type \\
  \hline
  Monomorphization
    & specializes polymorphic functions on their type arguments \\
  \hline
  Elimination of Equations
    & Translates multiple-equations definitions of functions into
      a single nested pattern matching \\
  \hline
  Polarization
    & specialize some predicates into a version used in positive position
      and one used in negative position \\
  \hline
  Unrolling
    & Add a decreasing argument to non-well-founded predicates \\
  \hline
  Skolemization
    & Introduce Skolem symbols for term variables \\
  \hline
  Elimination of Predicates
    & Changes a multi-clauses (co)inductive predicate definition
      into a single recursive function \\
  \hline
  Recursion Elimination
    & The encoding from Section~\ref{sec:encoding} \\
  \hline
  Elimination of Match
    & Transforms pattern-matching expressions into tests and selectors \\
  \hline
  CVC4 Invocation 
    & Call CVC4 to obtain a model \\
  \hline
\end{tabular}
%}}}
\caption{Pipeline of Transformations in Nunchaku}
\label{fig:pipeline}
\end{figure}

TODO: similar to institutions \cite{goguen-meseguer} (?)

\subsection{Decoding of Recursive Functions}

Nunchaku is designed for finding counter-models to formulas a user thinks
are theorems. Decoding models from CVC4 to hide any trace of the successive
encodings is therefore vital for readability. Each step
of the pipeline only has to concern itself with removing its changes to
the signature and logic; our encoding of recursive functions is no exceptions.

TODO explain how:
- domains of proj. functions are computed
- domains of original function is deduced
- decision tree for original function is built
- note how junk is removed in step 3 (but undefined "..." as last case)

\subsection{Support for Higher-Order Functions}

TODO:
- currying with explicit application symbol --> function arguments are now
  just constants
- apply encoding on application symbol
- gotcha: one function can be used with several arities.

- optim: specialization (similar to monomorphization)
